---
title: "Real-World Example: Hurricane Tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-World Example: Hurricane Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates TRACLUS on geographic trajectory data using
Atlantic hurricane tracks in HURDAT2 format. The bundled dataset
contains **826 named Atlantic storms from 1950 to 2004** with over
22,000 track points. No internet connection is required.

## Reading HURDAT2 data

The `tc_read_hurdat2()` function parses NOAA's HURDAT2 Best-Track format,
converting West longitudes to negative values and filtering storms with
fewer than a minimum number of track points:

```{r read-hurdat2}
library(TRACLUS)

filepath <- system.file("extdata", "hurdat2_1950_2004.txt",
  package = "TRACLUS"
)
storms <- tc_read_hurdat2(filepath, min_points = 3)
cat("Storms:", length(unique(storms$storm_id)), "\n")
cat("Points:", nrow(storms), "\n")
head(storms)
```

## Working with a subset

The full dataset contains hundreds of storms. For demonstration and
interactive exploration, it is practical to work with a subset.
Use `min_points` to select only storms with long tracks (which tend
to be the most significant hurricanes):

```{r subset}
# Select only storms with >= 80 track points (long-lived storms)
long_storms <- tc_read_hurdat2(filepath, min_points = 80)
cat("Long-lived storms:", length(unique(long_storms$storm_id)), "\n")
cat("Points:", nrow(long_storms), "\n")
```

## Creating trajectory objects

For geographic data, set `coord_type = "geographic"`. The x-column must
contain longitude and the y-column must contain latitude:

```{r trajectories}
trj <- tc_trajectories(long_storms,
  traj_id = "storm_id",
  x = "lon", y = "lat",
  coord_type = "geographic"
)
trj
```

```{r plot-trj}
plot(trj)
```

## Full pipeline

With geographic data, distances are in **metres**. TRACLUS offers three
methods for geographic data (set via the `method` argument in
`tc_trajectories()`):

- `"projected"` *(default)*: Equirectangular projection to metres, then
  euclidean distances. 5–10× faster than haversine with < 2 % error for
  regional datasets such as Atlantic hurricane tracks. Recommended for most
  use cases.
- `"haversine"`: Exact great-circle distances. Use for global datasets or
  when maximum accuracy is required.
- `"euclidean"`: Raw degree values treated as Cartesian. Only for paper
  replication.

A reasonable `eps` for Atlantic hurricanes is in the range of hundreds of
kilometres:

```{r pipeline}
# method = "projected" is the default — fast and accurate for regional data
result <- tc_traclus(trj, eps = 500000, min_lns = 3)
result
```

```{r plot-result}
plot(result)
```

```{r summary}
summary(result)
```

## Interactive Leaflet map

If the `leaflet` package is installed, `tc_leaflet()` creates an
interactive map with multiple tile layers and hover labels:

```{r leaflet, eval = requireNamespace("leaflet", quietly = TRUE)}
tc_leaflet(result)
```

The map supports three base layers (CartoDB Positron, OpenStreetMap,
Esri World Imagery) and can be further customised via standard
`leaflet` piping.

## Step-by-step with inspection

For more control, run each step individually:

```{r step-by-step}
parts <- tc_partition(trj)
parts

clust <- tc_cluster(parts, eps = 500000, min_lns = 3)
clust

repr <- tc_represent(clust)
repr
```

Each intermediate object can be plotted or mapped independently:

```{r plot-parts}
plot(parts)
```

```{r leaflet-parts, eval = requireNamespace("leaflet", quietly = TRUE)}
tc_leaflet(parts)
```

## Scaling to the full dataset

For production analyses with all 800+ storms, the clustering step
involves O(n²) pairwise distance comparisons and may take several
minutes. Consider:

- Using `method = "projected"` (the default) to reduce computation time
  by 5–10× compared to `"haversine"` with negligible accuracy loss.
- Using `tc_estimate_params()` on the full partitions to find good
  starting parameters.
- Pre-filtering storms by year range or geographic region.
- Starting with a higher `min_points` threshold to reduce the number
  of segments.
