---
title: "Geographic Distance Methods in TRACLUS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geographic Distance Methods in TRACLUS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

TRACLUS supports two coordinate systems:

- **Euclidean** (`coord_type = "euclidean"`): Standard 2D Cartesian
  coordinates. Distances computed using Euclidean geometry.
- **Geographic** (`coord_type = "geographic"`): Longitude/latitude
  coordinates on a sphere. Three distance methods are available via the
  `method` argument.

This vignette explains the three geographic distance methods and when to
use each one.

## Choosing a method

When `coord_type = "geographic"`, the `method` argument controls how
distances are measured:

| | `"haversine"` | `"projected"` | `"euclidean"` |
|---|---|---|---|
| **Accuracy** | Exact spherical | < 2 % error (regional) | Not metric |
| **Speed** | Slowest | ~ 5–10× faster | Fastest |
| **`eps` units** | Metres | Metres | Degree-units |
| **Sweep-line** | Equirectangular proj. | Equirectangular proj. | Direct coords |
| **Use case** | Global datasets, max accuracy | **Most geographic datasets** | Paper replication only |

**Rule of thumb:**

- Use `method = "projected"` (the default for geographic data) for regional
  datasets such as hurricane basins, city taxi data, or vessel AIS tracks.
  It is 5–10× faster than haversine with less than 2 % error.
- Use `method = "haversine"` when the dataset spans large portions of the
  globe (e.g., transoceanic flight paths) or when exact distances are
  required.
- Use `method = "euclidean"` only to replicate the original TRACLUS paper
  results exactly. Degree-unit distances are not physically meaningful.

```{r choosing-method}
library(TRACLUS)

filepath <- system.file("extdata", "hurdat2_1950_2004.txt",
  package = "TRACLUS"
)
storms <- tc_read_hurdat2(filepath, min_points = 80)

# Default: method = NULL resolves to "projected" for geographic data
trj_proj <- tc_trajectories(storms,
  traj_id = "storm_id",
  x = "lon", y = "lat",
  coord_type = "geographic"
)

# Explicit haversine for comparison
trj_hav <- tc_trajectories(storms,
  traj_id = "storm_id",
  x = "lon", y = "lat",
  coord_type = "geographic",
  method = "haversine"
)

cat("projected method:", trj_proj$method, "\n")
cat("haversine method:", trj_hav$method, "\n")
```

## How "projected" works

`method = "projected"` applies an **equirectangular projection** to convert
longitude/latitude coordinates to metres before computing distances:

$$x' = \text{lon} \times \cos(\bar{\varphi}) \times 111{,}320$$
$$y' = \text{lat} \times 111{,}320$$

where $\bar{\varphi}$ is the mean latitude of all trajectory points
(computed once during `tc_trajectories()`). Euclidean distances on these
projected coordinates are then used throughout the pipeline.

After the sweep-line generates representative waypoints in projected
coordinates, they are inverse-projected back to longitude/latitude.

This projection is accurate for datasets that cover a limited latitudinal
range (up to several thousand kilometres). The error grows at high
latitudes or for very large geographic extents, where `"haversine"` should
be preferred.

## Distance functions

When `coord_type = "geographic"`, the three distance components use
spherical geometry with `method = "haversine"`, or euclidean geometry
on projected coordinates with `method = "projected"`.

### Perpendicular distance ($d_\perp$) — haversine

Uses **cross-track distance**: the shortest distance from a point to
the great-circle arc defined by a line segment. Based on the formula:

$$d_{xt} = R \cdot \arcsin\left(\sin\left(\frac{d_{13}}{R}\right) \cdot \sin(\theta_{13} - \theta_{12})\right)$$
where $R = 6{,}371{,}000$ m (Earth's radius), $d_{13}$ is the haversine
distance from the segment start to the projection point, and
$\theta_{12}$, $\theta_{13}$ are initial bearings.

The perpendicular distance uses the Lehmer mean (order 2) of the two
cross-track distances, consistent with the euclidean definition.

### Parallel distance ($d_\parallel$) — haversine

Uses **along-track signed distance**: how far along the great-circle
arc a perpendicular foot falls. This provides the signed displacement
needed for the parallel distance computation.

### Angle distance ($d_\angle$)

Uses the **initial bearing** difference between two segments (haversine),
or the angle between direction vectors (projected/euclidean).
When the angle exceeds 90 degrees, the segment is reversed before
computing the angular penalty.

### Note on exported distance functions

The exported `tc_dist_*()` functions (`tc_dist_perpendicular()`,
`tc_dist_parallel()`, `tc_dist_angle()`, `tc_dist_segments()`) accept
`method = "haversine"` or `method = "euclidean"`. The `"projected"` method
is applied internally by the pipeline functions (`tc_partition()`,
`tc_cluster()`, `tc_represent()`) and is not exposed in the low-level
distance API.

### Example: comparing haversine distances

```{r distance-example}
# Two segments near the English Channel
si <- c(-0.1278, 51.5074) # London
ei <- c(2.3522, 48.8566) # Paris
sj <- c(0.5, 50.5) # mid-Channel
ej <- c(3.0, 49.5) # Belgian coast

# Spherical distances (metres)
d_perp_sph <- tc_dist_perpendicular(si, ei, sj, ej, method = "haversine")
d_par_sph <- tc_dist_parallel(si, ei, sj, ej, method = "haversine")
d_ang_sph <- tc_dist_angle(si, ei, sj, ej, method = "haversine")
d_tot_sph <- tc_dist_segments(si, ei, sj, ej, method = "haversine")

cat(sprintf("Perpendicular: %.0f m\n", d_perp_sph))
cat(sprintf("Parallel:      %.0f m\n", d_par_sph))
cat(sprintf("Angle:         %.0f m\n", d_ang_sph))
cat(sprintf("Total:         %.0f m\n", d_tot_sph))
```

## Partitioning

MDL-based partitioning uses haversine distances (`method = "haversine"`)
for both the $L(H)$ (hypothesis cost) and $L(D|H)$ (data-given-hypothesis
cost) terms. With `method = "projected"`, coordinates are first projected
to metres, and then euclidean MDL distances are applied — producing the
same partition structure with lower computational cost.

The partition points are the same characteristic points as in the
euclidean case, but distances are measured along great-circle arcs
(haversine) or in projected metres (projected).

## Clustering

Neighbourhood computation uses the distance method chosen at
`tc_trajectories()` time. The `eps` parameter is in **metres** for both
`"haversine"` and `"projected"`:

```{r geographic-workflow}
trj <- tc_trajectories(storms,
  traj_id = "storm_id",
  x = "lon", y = "lat",
  coord_type = "geographic"
) # method = "projected" default

parts <- tc_partition(trj)
clust <- tc_cluster(parts, eps = 500000, min_lns = 3)
clust
```

## Representation

The sweep-line algorithm operates in projected coordinates for both
`"haversine"` and `"projected"`. For `"haversine"` data, TRACLUS applies
the equirectangular projection internally just for the sweep-line step,
then inverse-projects the waypoints. For `"projected"` data, the
coordinates are already in projected metres throughout the pipeline.

```{r represent}
repr <- tc_represent(clust)
repr
plot(repr)
```

## Aspect ratio correction

When plotting geographic data with base R `plot()`, the aspect ratio
is corrected using:

$$\text{asp} = \frac{1}{\cos(\bar{\varphi})}$$

This compensates for the convergence of meridians at higher latitudes,
producing visually accurate proportions.

## Interactive maps

For geographic data, `tc_leaflet()` provides interactive Leaflet maps
that handle projection automatically:

```{r leaflet, eval = requireNamespace("leaflet", quietly = TRUE)}
tc_leaflet(repr)
```

Leaflet maps are only available for geographic data. Attempting to use
`tc_leaflet()` on euclidean data will produce an informative error.

## Anti-meridian limitation

**Known limitation:** Trajectories that cross the anti-meridian
(180°/−180° longitude) can produce incorrect distance calculations.
The haversine formula itself handles the wrap-around correctly, but
the coordinate arithmetic in partitioning, clustering, and
representation assumes a continuous longitude range. A jump from
179° to −179° (a 2° real-world gap) appears as a 358° discontinuity
in the coordinate space.

### Workaround

Shift all longitudes to the range [0°, 360°] before processing:

```r
# Shift to [0, 360] before creating trajectories
my_data$lon <- ifelse(my_data$lon < 0, my_data$lon + 360, my_data$lon)

trj <- tc_trajectories(my_data, traj_id = "id",
                       x = "lon", y = "lat",
                       coord_type = "geographic")

# Run the full pipeline in [0, 360] space
result <- tc_traclus(trj, eps = 500000, min_lns = 3)

# Shift representative waypoints back to [-180, 180] for display
# (tc_leaflet and most mapping tools expect [-180, 180])
```

### Limitation of the workaround

This shift works **only** when no trajectory crosses both the prime
meridian (0°) **and** the anti-meridian (180°). For transpacific routes
that span from, say, East Asia to the Americas, the shifted coordinates
may still be discontinuous. A robust solution would require segment-level
longitude unwrapping, which is not implemented in version 1. This is
documented honestly — for such datasets, users should consider
pre-processing their coordinates or splitting trajectories at the
anti-meridian.

## Key differences summary

| Aspect | Euclidean | Geographic / haversine | Geographic / projected |
|--------|-----------|------------------------|------------------------|
| Input coordinates | x, y (any units) | longitude, latitude (degrees) | longitude, latitude (degrees) |
| Distance method | Euclidean | Haversine (spherical) | Equirectangular + Euclidean |
| eps units | Coordinate units | Metres | Metres |
| $d_\perp$ | Perpendicular foot distance | Cross-track distance | Projected perpendicular |
| $d_\parallel$ | Along-segment projection | Along-track signed distance | Projected parallel |
| $d_\angle$ | sin(angle) × segment length | sin(bearing diff) × arc length | sin(angle) × proj. length |
| Sweep-line | Direct coordinates | Equirectangular projection | Already in projected metres |
| Plot asp | 1 | 1/cos(mean latitude) | 1/cos(mean latitude) |
| Leaflet | Not available | Available | Available |
| Speed | Fast | Slowest | ~ 5–10× faster than haversine |
| Accuracy | N/A (no real-world metric) | Exact | < 2 % error (regional) |
