---
title: "Parameter Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The TRACLUS algorithm has two critical parameters:

- **`eps`** — the neighbourhood radius for DBSCAN clustering
- **`min_lns`** — the minimum number of neighbouring segments to form a
  cluster

Choosing appropriate values is essential for meaningful results. This
vignette explains the parameters and demonstrates the entropy-based
estimation heuristic.

## Understanding eps

`eps` defines the maximum weighted distance between two line segments
for them to be considered neighbours. The weighted distance is:

$$d(L_i, L_j) = w_\perp \cdot d_\perp + w_\parallel \cdot d_\parallel + w_\angle \cdot d_\angle$$

where the three components capture perpendicular offset, parallel
displacement, and angular difference respectively.

- **Euclidean data:** `eps` is in the same coordinate units as the data.
- **Geographic data:** `eps` is in **metres** (`method = "haversine"` or
  `"projected"`).

### Effect of eps

- **Too small:** Every segment is noise; no clusters are found.
- **Too large:** All segments merge into a single cluster, losing
  structure.
- **Just right:** Meaningful clusters emerge that correspond to common
  trajectory patterns.

```{r eps-demo}
library(TRACLUS)

trj <- tc_trajectories(traclus_toy, traj_id = "traj_id",
                       x = "x", y = "y", coord_type = "euclidean")
parts <- tc_partition(trj)

# Too small: 0 clusters
c1 <- suppressWarnings(tc_cluster(parts, eps = 1, min_lns = 3))
cat("eps = 1:  ", c1$n_clusters, "clusters\n")

# Reasonable: meaningful clusters
c2 <- suppressWarnings(tc_cluster(parts, eps = 25, min_lns = 3))
cat("eps = 25: ", c2$n_clusters, "clusters\n")

# Too large: everything in one cluster
c3 <- suppressWarnings(tc_cluster(parts, eps = 500, min_lns = 3))
cat("eps = 500:", c3$n_clusters, "clusters\n")
```

## Understanding min_lns

`min_lns` sets the minimum number of neighbouring segments a core
segment must have for DBSCAN expansion. Additionally, after clustering,
any cluster whose member segments come from fewer than `min_lns` distinct
trajectories is removed (the **cardinality check** from the paper).

- **Too small (e.g., 1):** Almost every segment forms its own cluster.
- **Too large:** Only very dense regions form clusters; most segments
  become noise.

### Why min_lns < 3 can produce misleading results

`min_lns` also controls the sweep-line density threshold in
representative generation (Phase 3). The sweep-line generates a waypoint
only when enough segments cross the current position. However, when
consecutive segments from the **same trajectory** share an endpoint
(a characteristic point), both segments are momentarily counted at that
position due to the entry-before-exit event processing. With
`min_lns = 2`, this momentary peak of 2 is enough to create a waypoint —
even though both segments come from a single trajectory.

Three or more consecutive segments from one trajectory create 2+
waypoints at their shared endpoints, producing a representative that
simply traces that trajectory's path instead of capturing a common
sub-trajectory from multiple sources.

**Our fix:** The sweep-line additionally checks that at least 2 distinct
trajectories are represented among the active segments before generating
a waypoint. This eliminates the degenerate case without affecting
results for `min_lns >= 3`.

The original TRACLUS paper (Lee, Han & Whang 2007) does not include this
check because all experiments use MinLns = 6--9 (Section 4.4), where the
edge case cannot occur. Our extension is consistent with the paper's
Phase 2 cardinality check, which also requires trajectory diversity.

**Recommendation:** Use `min_lns >= 3`, or better yet, use the
data-driven estimate from `tc_estimate_params()` which typically suggests
values of 5 or higher.

## Automatic parameter estimation

`tc_estimate_params()` provides a data-driven starting point using
the entropy heuristic from Section 4.4 of the TRACLUS paper:

1. Compute pairwise distances for a sample of segments.
2. For each candidate `eps`, count how many neighbours each segment has.
3. Compute the Shannon entropy of the neighbourhood-size distribution.
4. Select the `eps` that **minimises** entropy (sharpest clustering
   structure).
5. Set `min_lns = ceiling(mean neighbourhood size at optimal eps) + 1`.

```{r estimate}
set.seed(42)
est <- tc_estimate_params(parts)
est
```

```{r plot-estimate}
plot(est)
```

The red circle marks the optimal `eps` value. Use these estimates as a
starting point, then refine based on domain knowledge and visual
inspection.

## Using estimated parameters

```{r use-estimate}
result <- suppressWarnings(
  tc_traclus(trj, eps = est$eps, min_lns = est$min_lns)
)
result
plot(result)
```

## Distance weights

The three distance components can be weighted differently using
`w_perp`, `w_par`, and `w_angle`:

```{r weights}
# Emphasise angular similarity
c_angle <- suppressWarnings(
  tc_cluster(parts, eps = 25, min_lns = 3,
             w_perp = 0.5, w_par = 0.5, w_angle = 2)
)
cat("Angle-heavy: ", c_angle$n_clusters, "clusters\n")

# Emphasise perpendicular distance
c_perp <- suppressWarnings(
  tc_cluster(parts, eps = 25, min_lns = 3,
             w_perp = 2, w_par = 0.5, w_angle = 0.5)
)
cat("Perp-heavy:  ", c_perp$n_clusters, "clusters\n")
```

## The gamma parameter

The `gamma` parameter in `tc_represent()` (and `tc_traclus()`) controls
the minimum spacing between waypoints in the representative trajectory.
The default value of 1 works well for most cases:

- **Smaller gamma:** More waypoints, higher fidelity to the underlying
  segments.
- **Larger gamma:** Fewer waypoints, smoother representative.

```{r gamma}
repr1 <- tc_represent(c2, gamma = 1)
repr5 <- tc_represent(c2, gamma = 5)

cat("gamma = 1:", nrow(repr1$representatives), "total waypoints\n")
cat("gamma = 5:", nrow(repr5$representatives), "total waypoints\n")
```

## Guidelines

| Parameter | Euclidean | Geographic |
|-----------|-----------|------------|
| `eps` | Same units as coordinates | Metres (e.g., 100000 = 100 km) |
| `min_lns` | 3--5 for small datasets | 3+ (use `tc_estimate_params()`) |
| `gamma` | 1 (default) usually fine | 1 (default) usually fine |
| Weights | Equal (default) unless domain-specific | Equal (default) |
| `method` | N/A | `"projected"` (default, fast) or `"haversine"` (exact) |

**Practical workflow:**

1. Run `tc_estimate_params()` for a data-driven starting point.
2. Try the suggested values with `tc_traclus()`.
3. Inspect `plot()` and `summary()` output.
4. Adjust `eps` up (fewer clusters) or down (more clusters) as needed.
5. Fine-tune `min_lns` if noise/cluster balance is off.
