---
title: "Getting Started with TRACLUS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with TRACLUS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **TRACLUS** package implements the trajectory clustering algorithm by
Lee, Han & Whang (SIGMOD 2007). The algorithm works in three phases:

1. **Partition** trajectories into line segments using Minimum Description
   Length (MDL).
2. **Cluster** segments via DBSCAN density-based spatial clustering.
3. **Represent** each cluster with a representative trajectory using a
   sweep-line algorithm.

This vignette walks through the complete workflow using the built-in toy
dataset.

## Setup

```{r setup}
library(TRACLUS)
```

## Step 1: Load and prepare data

TRACLUS expects a data.frame with columns for trajectory ID, x-coordinate,
and y-coordinate. The built-in `traclus_toy` dataset contains 6 euclidean
trajectories:

```{r load-data}
head(traclus_toy)

trj <- tc_trajectories(traclus_toy, traj_id = "traj_id",
                       x = "x", y = "y", coord_type = "euclidean")
trj
```

Visualise the raw trajectories:

```{r plot-trajectories}
plot(trj)
```

## Step 2: Partition trajectories

MDL-based partitioning splits each trajectory at characteristic points
where the trajectory changes direction significantly:

```{r partition}
parts <- tc_partition(trj)
parts
```

```{r plot-partitions}
plot(parts)
```

The black X markers show the characteristic points selected by the
MDL criterion. Each trajectory is now represented as a sequence of
line segments.

## Step 3: Cluster segments

DBSCAN groups similar line segments into clusters. Two key parameters
control the result:

- **`eps`**: Maximum distance between segments to be considered neighbours.
- **`min_lns`**: Minimum number of neighbouring segments to form a cluster.

```{r cluster}
clust <- tc_cluster(parts, eps = 25, min_lns = 3)
clust
```

```{r plot-clusters}
plot(clust)
```

Segments that do not belong to any cluster are classified as **noise**
(shown as dashed grey lines).

## Step 4: Generate representative trajectories

The sweep-line algorithm creates a representative trajectory for each
cluster by averaging the contributing segments:

```{r represent}
repr <- tc_represent(clust)
repr
```

```{r plot-representatives}
plot(repr)
```

Use `show_clusters = TRUE` to see the cluster segments in colour
alongside the representatives:

```{r plot-repr-clusters}
plot(repr, show_clusters = TRUE)
```

## One-call pipeline

For convenience, `tc_traclus()` runs the entire pipeline in a single call:

```{r traclus-wrapper}
result <- tc_traclus(trj, eps = 25, min_lns = 3)
result
```

```{r plot-traclus}
plot(result)
```

## Inspecting results

Every object has `print()`, `summary()`, and `plot()` methods:
```{r summary}
summary(result)
```

## Trial-and-error: re-clustering with different parameters

A key design feature of TRACLUS is that each phase returns a **new
object** without modifying its input. This means you can re-use the
same `tc_partitions` object to try different `eps` and `min_lns`
values — no need to re-partition every time:
```{r trial-and-error}
# Partition once
parts <- tc_partition(trj)

# Try different clustering parameters on the same partitions
clust_tight  <- tc_cluster(parts, eps = 15, min_lns = 3)
clust_medium <- tc_cluster(parts, eps = 25, min_lns = 3)
clust_wide   <- tc_cluster(parts, eps = 40, min_lns = 2)

cat(sprintf("eps=15: %d cluster(s), %.0f%% noise\n",
            clust_tight$n_clusters,
            100 * clust_tight$n_noise / nrow(clust_tight$segments)))
cat(sprintf("eps=25: %d cluster(s), %.0f%% noise\n",
            clust_medium$n_clusters,
            100 * clust_medium$n_noise / nrow(clust_medium$segments)))
cat(sprintf("eps=40: %d cluster(s), %.0f%% noise\n",
            clust_wide$n_clusters,
            100 * clust_wide$n_noise / nrow(clust_wide$segments)))
```

This makes parameter exploration fast and convenient. See
`vignette("TRACLUS-parameter-guide")` for systematic guidance on
choosing good values.

## Which `coord_type` do I need?

When creating trajectories with `tc_trajectories()`, the `coord_type`
argument tells TRACLUS how to measure distances:
| Your data | `coord_type` | `eps` units |
|-----------|--------------|-------------|
| Synthetic / simulation (arbitrary x, y) | `"euclidean"` | Same as your coordinates |
| GPS tracks, ship routes, flight paths (lon, lat in degrees) | `"geographic"` | **Metres** |
| OpenStreetMap or any WGS-84 source | `"geographic"` | **Metres** |

**Coordinate convention for geographic data:** `x` = longitude (−180
to 180), `y` = latitude (−90 to 90). This follows the mathematical
convention (x = horizontal axis). If your data has latitude first,
swap the column assignments:

```r
# Latitude-first data? Just swap x and y:
trj <- tc_trajectories(df, traj_id = "id",
                       x = "longitude", y = "latitude",
                       coord_type = "geographic")
```

TRACLUS will warn you if it detects that x and y might be swapped
(e.g., all x values in [−90, 90] while y values are outside that
range).

For geographic data, you can also control *how* distances are computed
with the optional `method` argument:

- `method = "projected"` *(default)*: 5–10× faster than haversine with
  < 2 % error. Recommended for most regional datasets.
- `method = "haversine"`: Exact great-circle distances. Use for global
  datasets or when maximum accuracy is needed.

See `vignette("TRACLUS-spherical-geometry")` for a full comparison of
the three available methods.

## Next steps

- See `vignette("TRACLUS-real-world")` for an example with real hurricane
  track data.
- See `vignette("TRACLUS-parameter-guide")` for guidance on choosing
  `eps` and `min_lns`.
- See `vignette("TRACLUS-spherical-geometry")` for details on geographic
  (spherical) distance calculations.
